Project Report

### 2022-01-11 Gandhi_Project_Report/draft3.tex
Submitted to Gandhi for review. Still lots to be done

% \documentclass[a4paper,12pt,twoside]{book}
\documentclass[a4paper,12pt,twoside]{report}
\pagestyle{headings}
\usepackage[utf8]{inputenc}
\usepackage{filecontents}
\usepackage[many]{tcolorbox}
\usepackage{graphicx}
\usepackage{harpoon}
%\usepackage{apacite} 
\usepackage{xcolor}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mdframed}
%\usepackage[authoryear]{natbib}
\usepackage{hyperref, url}
\usepackage[show]{ed}
\usepackage{subfigure}
\usepackage[ntheorem]{empheq} 
\usepackage{amsthm, amssymb, amsfonts, latexsym}
% \usepackage{pst-all}

@@

\title{Project Thesis/Report}

@@

Examples in the literature attempting to forecast data from such systems have tried a number of different approaches with differing degrees of success. 
When we have complete access to states of the system, an ordinary differential equation can be obtained from data (Citations to Sindy and \cite{small2002modeling,xu2006modeling}) wherein one could approximate the vector field by a library of functions to obtain interpretable models.

Recent partially interpretable models available in the literature have been based on the Koopman operator (e.g.,\cite{koopman1932dynamical,budivsic2012applied}) to  employ observables mapping the data onto a higher-dimensional space. This then makes the dynamics in the higher-dimensional space more amenable for approximation by a linear transformation. Such methods, obviously, do not guarantee an exact reconstructions for nonlinear models, and in practice provide poor long-term consistency for a large class of systems. (Add citation)

The non-interpretable models include the delay embedding and the machine learning algorithms. One could learn a system conjugate to the underlying system by applying the Takens delay embedding \cite{takens1981detecting} when one has ``good" observations from the system; this learnt system could then be used for forecast the observed data.  Takens delay embedding theorem \cite{takens1981detecting} and its various generalisations (e.g., \cite{sauer1991embedology, stark1999delay, gutman2018embedding, gutman2018embedding}) establishes the learnability of a system constructed by concatenating sufficiently large previous time-series observations of a dynamical system into a vector (called delay coordinates). This then establishes the existence of a map on the space of delay coordinates equivalent (or topologically conjugate) to the underlying map from which the observed time-series was first obtained. Although topological conjugacy guarantees an alternate representation of the underlying system, the quality of this representation still depends on numerous parameters, making the comprehension of the dynamics rather unreliable (e.g., \cite{principe1992prediction}). One reason for this fragility is that the embedded attractor in the reconstruction space not always an attractor of the map that is learnt in the reconstruction space, despite  unquestionably being an invariant set. When the embedded object is not \textbf{(definitely/always?)} an attractor (as explained in Chapter 3) it can cause predictions to fail.

@@

This thesis deals with the implementation and analysis of non-interpretable models that can guarantee \textbf{exact/accurate} reconstruction. The project work concerns the study and implementation of a method (Citation to paper with Adriaan) that incorporates learning a function by mapping the data on to a higher dimensional space using what is called a driven dynamical system (See Chapter 4). With a clear understanding of how the data is mapped onto the higher dimensional space, the method then permits the learning of a dynamical system topologically conjugate to that of the underlying system. Instead of linear regression as in the training of echo state networks, deep learning methods are employed to learn the correct function. With slight modifications to the implementation in the paper (with Adriaan), we show that one can construct accurate non-interpretable models with the ability to reconstruct attractors from more hard-to-forecast systems like the double pendulum. (The forecasting of the time-series from a double pendulum has not been reported before.) Moreover, we also demonstrate that long-term statistical consistency is preserved.

By solidifying the mathematical underpinnings of our theory, we hope to guarantee the ability to construct models with predictive power ranging from molecular biology to neuroscience. in the near future. (We can modify this sentence when the report is completed.)

The report is organised as follows: 
\newline In Chapter~\ref{ch2}, we recall the definition of a discrete-time dynamical system, how a discrete-system arises from a flow of an ODE and then proceed to define the inverse limit space and topological conjugacy of autonomous systems. 
\newline In Chapter~\ref{ch3}, we introduce the problem of forecasting dynamical systems, state the Takens delay embedding theorem and discuss various issues faced while forecasting. 
\newline In Section 4, a driven dynamical system is defined and discuss the properties of a specific class of driven dynamical systems that we make use of in this thesis/report/paper. 
\newline Finally in Section 5 we show the implementation of these forecasting methods, and conclusions are provided in Section 6.

\chapter{Discrete-time Dynamical Systems} \label{ch2}

@@

Of course, discrete-time dynamical systems need not always arise through a differential equation, and once again we may consider the field of ecology where discrete dynamical systems are often directly derived or assumed. \textbf{Cite.}
There is a school of thought that advocates discrete-time dynamical systems to be more natural for modelling real-world observations than differential equations and we refer the interested reader to \cite{saber2010introduction}.

@@

\ednote{For let $B$ be an attractor of U. From above, we know that B is itself invariant, is asymptotically stable and has Lyapunov stability. Now suppose there is some attractor $C\subsetneq{U}=B$, so there is a $b\in{U}$ which is not in $C$. ........  M: Have changed the example}

@@

  In a more rigourous fashion ....
  \ednote{B:Should I be more rigourous here? }

@@

\ednote{M: Saying $\phi$ is then a semi-conjugacy between $T$ and $S$ could be ambiguous as it does not specify which system is on the top in the commutativity. So please edit accordingly. B: Does defining explicitly the domain and codomain of $phi$ make it clear enough? }
When $S$ is conjugate to $T$, the dynamics of the two systems are in some way ''dynamically equivalent''. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$ \cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is also common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (e.g., \cite{de2013elements}) is a larger system capturing all of the important dynamics of its factor.

@@

First we establish the method of Takens delay embedding. 
In this method, one considers a discrete-time dynamical system defined on a 'nice' space (a smooth manifold)\ednote{Should this be further explained?} that can be obtained as time-$K$ map of a continuous time-dynamical system \ednote{B: Should one add here: "These notions will be made formal below"?}. 
We make observations from such a system, i.e., we observe the evolution of $\theta(w_0)$ (where $w_0$ represents the initial state) i.e., we observe  a finite set of values of $u_n :=  \theta(w_n) = \theta(Tw_{n-1})$. The sequence $\{u_n\}$ represents a scalar time-series and intuitively $\theta$ may be thought to represent a probe inserted into a larger system which is itself only measuring/extracting a small part of the greater system state at time $t=n$. 
Consider for example a thermometer erected to measure the ambient temperature in a local village. This measurement function, the thermometer, is capturing only a single aspect, the temperature, of a much grander dynamical system entailing the present weather of the surrounding area. Even more than that though- it is measuring a miniscule part of the global weather system.

@@

%Some explanation is beneficial. Takens establishes a delay-coordinate map $\Ftheta$ defined in \ref{eqn_takens} for $T$ the flow (\ref{defn_flow}) defined by the update equation  $Tu_n=u_{n+1}$ and $theta$ our measurement function. When confining a dynamical system to the manifold U (in which the underlying system is contained), Takens showed that if certain smoothness conditions are satisfied on $T$ and $\theta$, then the delay-coordinate map $\Ftheta$ embeds $U$ in the reconstruction space $R^{2d+1}$ \textbf{we haven't used the term reconstruction space yet, so this is falling out of the blue}, for almost every choice of measurement function $\theta$, the observable. By almost every choice here, we mean that we exclude some observations which do not give any information. To see a more precise discussion, see \textbf{xxx} for we do not concern ourselves with a further elaboration on this thesis/report.

@@

\begin{figure}[ht]
  \includegraphics[scale=0.25]{_takensmap.png}
  \centering
  \captionof{figure}{Schematic of Takens Embedding Theorem's usage and consequence of how iterates \ldots }
  \label{fig:takensmap}
\end{figure}
\ednote{M: Here, and throughout if you reproduce a figure, it must be stated where it was reproduced from. B: How does one write that a figure is reproduced? Also, should $\tilde{F}$ be left out for now?}
\ednote{B: Could I perhaps have the original figure? This is only a screenshot used as a temporary placeholder.}

@@

%Thus information about U can be retained in the time series' (or observation's) output. By preserving the topology on the manifold $U$ in the reconstruction space $X$ (\textbf{not discussed before}), we are also guaranteed the preservation of topological invariants of the manifold, of which dimensionality is one such invariant. (We note here that dimensionality will again be considered later in this project/report).

@@

It is a fact that Takens' Embedding Theorem is both a powerful result and that it provides compelling reason to believe that one could conceivably reconstruct accurately a system conjugate to the underlying system. Nonetheless, it does present some serious practical limitations:
\vspace{-5mm}
\begin{enumerate}
\item Even supposing that we could indeed find $F_\theta$ , our \emph{approximation} of  $F_\theta$ is a map from a larger set $\mathbb{R}^{2d+1}$ containing the embedded attractor. There are, however, no theoretical guarantees that $F_\theta$ will retain  $V=\Phi_{2d,\theta}(W)$  as an attractor although $W$ could itself be an attractor.
\item Takens theorem is stated for noiseless observations. Due to noise $\epsilon_n$ the delay vector $\Phi_{2d,\theta}(w_n) + \epsilon_n$  may lie outside $V$.Furthermore, due to the chaotic nature of the underlying system (i.e. the fact that it has SIDC), the evolution of $\Phi_{2d,\theta}(w_n) + \epsilon_n$ under the map $F_\theta$ could move out of $V$ completely. This problem can be overcome  by using a driven dynamical system with some properties, and we discuss this in Chapter *. 
%  \item It is an unfortunate case that the functional complexity (i.e., the relative number of oscillatory graphs \cite{manjunath2021universal}) as well as stability of the embedding \Ftheta{} (\textbf{cite?})depends on the choice of observable $\theta$. This means that we may even learn a map that complicates our learning task and in so doing adds obstacles to the road to learning a conjugate map with predictive power. \ednote{Am I talking nonsense here? M: Please avoid this point and delete it}
\end{enumerate}

@@

\textbf{Continue writing here}

\begin{figure}[ht]
  \includegraphics[scale=0.3]{_learningfailure.png}
  \centering
  \label{fig:supp_learning_failure}
  \caption{Illustration of a system having failed to learn the map}
\end{figure}

\textbf{Explanation continued here}

\begin{figure}[ht]
  \includegraphics[scale=0.4]{_dp_dying.eps}
  \centering
  \label{fig:damped_pendulum}
\caption{A damped pendulum}
\end{figure}

\textbf{Continue explaining here}

@@

\section{Next section? \ednote{I feel a break is necessary here, but would hear from you what your thoughts are.}}

@@

\begin{Theorem}
  For a compactly driven system, a causal mapping $H$ exists if and only if $g$ has the USP. 
\end{Theorem}
\begin{proof}
  See \cite[Th.3]{manjunath2013echo}.
\end{proof}

\textbf{We direct our attention to the function $H_2(\overline{u}):=(h(r\overline{u}, h(\overline{u})))$. }

@@

The driven system $g$ can induce an embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ as follows: If $H:\overleftarrow{U}\to\overleftarrow{X}_U$ is also injective (in addition to being surjective, and hence bijective), it becomes the embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ induced by the driven system $g$ and we refer to $H$ as a \emph{causal embedding} \ednote{M : Causal embedding a dynamical system is different from the definition of causal embedding, so please edit/delete this paragraph and the next theorem and also in the remaining part till Section 1.2}.

@@

When $g$ has the USP, the diagram below illustrates the operation of the the mappings $h$ and $H$. The mapping $h:\overleftarrow{U}\to{X_U}$ is an observable as discussed in an above section (\ref{TakensEmbedding}).

@@

% We have not as of yet treated the UAP properly, and establish the necessary result here (without proof) so as not to break up the completeness of theorem.

\section {Choosing the driven system $g$}

@@

When a causal embedding $H$ exists for the driven system $g$, one can map an arbitrary input ${u}$ onto the solution space $X$ without additional distortion or information-loss \textbf{(Cite.')}.
When an embedding is established, the question of possible additional complexity in the solution is removed by guaranteeing that, since the systems are conjugate (semi-conjugate, \textbf{(refer)}), $g$ does not add any (some) complexity to the system.

@@

\begin{Definition}
  [\bf Uniform Attraction Property]\label{Dfn_UAP}\rm
  A driven system $g$ has the Uniform Attraction Property (UAP) if, regardless of starting position, all trajectories \textbf{(of what?)} converge to a single trajectory as time flows forward. 
  This trajectory is also the unique solution sequence $x$ to the input sequence $u$ as mentioned above \ref{here_UAP_firstestablished}.
\end{Definition}

@@

%Change the theorem above so that th.4 is split up into two parts.

\begin{figure}[ht]
  %\includegraphics[scale=0.4]{memloss_conttime.png}
  \centering
\captionof{figure}{Approaching the true solution if g has the UAP}
\label{fig:memloss_conttime.png} 
\end{figure}

@@

This also solves the problem of perturbation/noise introduced by the observable or measurement function. Recall from before that an observable is inherently a map that discretises the underlying continuous-time system. Moreover, measurement and device error introduce mistakes. Since we are working with a chaotic system displaying sensitive dependence on initial conditions, these small errors could potentially send the trajectory into a completely different attractor. 
In the presence of the UAP, however, small measurement errors introduced into the system do not pose the same danger as before since the trajectories will eventually tend towards the actual solution.
\ednote{(B: Complete enough?)}
% \textbf{From the perspective of perturbation of an autonomous dynamical system, if the resulting nonautonomous system has the USP, then the temporal structure of noise relates to the dynamics on a nonautonomous uniform attractor through a topological semi-conjugacy or a conjugacy.}

@@

\textbf{NB NB NB - Causal Embedding Theorem! - Th.4}

@@

\begin{figure}[ht]
  \includegraphics[scale=0.3]{_gammatemp.png}
  \centering
  \label{fig:temp_Gamma}
\end{figure}

@@

Informally, we may consider input-related stability as concerned with the question of whether or not small variations in input would result in small responses.   
If two inputs $u$ and $u_noisy$ are close in the product topology, then their tails could differ greatly due to the metric's insensitivity to the differences in sequence's tails.
 However if the system has the USP then the function $h$ is continuous and so $h(u), h(u_noisy)$ remain near – maintaining a measure of robustness to input and measurement noise.\cite{manjunath2021universal}.
In a similar fashion, parameter-related stability is obtained thanks to a result \cite[Lemma 3.2]{manjunath2020stability} relating this form of stability with the ESP (which is here equivalent to the USP).

@@

\begin{figure}[ht]
  \includegraphics[scale=0.4]{_AutovsDriven.eps}
  \centering
  \label{fig:learningFailure}
  \captionof{figure}{Illustrating empirically the importance of global dissipativity. Plotted below the three principal components of randomly initiated trajectories of a learnt version of $G_T$ versus the trajectory $(x_{n-1}, x_n)$(grey) of driven stated obtained from the Thomas Attractor's data (see below). }
\end{figure}

@@

We make use of a Recurrent Neural Network (RNN) in our implementation. RNN's are widely used in the literature for discrete-time processing and are considered to be discrete-time version of continuous dynamical systems modelling a network of neurons. (\textbf{Cite.})
A complete discussion of RNN's is outside the scope of this thesis, so we present only the details necessary to the current discussion. For information, (\textbf{Cite, Cite, Cite}) may be consulted. We implement our RNN in the \emph{Python} programming language and making use of the \emph{Scipy} and \emph{Tensorflow} libraries.

If the RNN we will be using in the next step has $N$ neurons, then $A$ and $B$ are each $N\times{N}$ matrices. Inputs $u_n$ of dimension $K < N$ are padded by zeroes to obtain an input of dimension $N$. As we do not need the entire left-infinite history of input, an arbitrary initialisation of the system with initial values followed by the driving of the system with the function $g$ means that the generated values will approach the actual solution’s elements in a uniform manner thanks to the UAP. \textbf{(Cite.)}

@@

\section{Changes in this thesis/project/report to previous work of~\cite{manjunath2021universal}}

We made several changes which gave improved results:
\vspace{-8mm}
\begin{enumerate}
\item Using Delay Coordinates (DC) when driving the system vs. using Single/Multi Delay Dynamics (SDD/MDD) after driving (and before training). Explain/show why these two methods are equivalent. \ednote{Would MDD  be not computationally intensive ? So may be you just explain if you can’t show it.}
\item Sparse matrix implementation. Slight improvement in some scenarios, but not fantastic. \ednote{Did it not help you a great deal with the DP forecasting?}

(Possible paragraph here/later on implementing sparsity of A/B: Generate sparse matrix that is also well-conditioned and correctly scaled matrix.)
\item Density of matrices A/B (separate from sparse matrix implementation) \ednote{Is it the probability distributions of the entries in A and B?}
\end{enumerate}

@@

The Double Pendulum consists of two pendulums fastened to one another such that the system moves as a whole. The first (or higher pendulum) is attached to a fixed point while the second is attached to the endpoint of the first. (See below)

\begin{figure}[ht] \label{Fig_Gamma}
  \includegraphics[scale=0.25]{_dbl_Pend.png}
  \centering
  \label{fig:conjugacy_with_Gamma}
\end{figure}

>>>>

\documentclass[a4paper,12pt,twoside]{report}
% \documentclass[a4paper,12pt,twoside]{book}
\pagestyle{headings}
\usepackage[utf8]{inputenc}
\usepackage{filecontents}
\usepackage[many]{tcolorbox}
\usepackage{graphicx}
\usepackage{harpoon}
%\usepackage{apacite} 
\usepackage{xcolor}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mdframed}
%\usepackage[authoryear]{natbib}
\usepackage{hyperref, url}
\usepackage[show]{ed}
\usepackage{subfigure}
\usepackage[ntheorem]{empheq} 
\usepackage{amsthm, amssymb, amsfonts, latexsym}
% \usepackage{pst-all}

@@

\title{Project Report}

@@

Examples in the literature attempting to forecast data from such systems have tried a number of different approaches with differing degrees of success. 
When we have complete access to states of the system, an ordinary differential equation can be obtained from data (\cite{brunton2016discovering, champion2019data} and \cite{small2002modeling,xu2006modeling}) wherein one could approximate the vector field by a library of functions to obtain interpretable models.

Recent partially interpretable models available in the literature have been based on the Koopman operator (e.g.,\cite{koopman1932dynamical,budivsic2012applied}) to  employ observables mapping the data onto a higher-dimensional space. This then makes the dynamics in the higher-dimensional space more amenable for approximation by a linear transformation. Such methods, obviously, do not guarantee an exact reconstructions for nonlinear models, and in practice provide poor long-term consistency for a large class of systems. \textbf{(Citation)}

The non-interpretable models include the delay embedding and the machine learning algorithms. One could learn a system conjugate to the underlying system by applying the Takens delay embedding \cite{takens1981detecting} when one has ``good" observations from the system; this learnt system could then be used for forecast the observed data.  Takens delay embedding theorem \cite{takens1981detecting} and its various generalisations (e.g., \cite{sauer1991embedology, stark1999delay, gutman2018embedding}) establishes the learnability of a system constructed by concatenating sufficiently large previous time-series observations of a dynamical system into a vector (called delay coordinates). This then establishes the existence of a map on the space of delay coordinates equivalent (or topologically conjugate) to the underlying map from which the observed time-series was first obtained. Although topological conjugacy guarantees an alternate representation of the underlying system, the quality of this representation still depends on numerous parameters, making the comprehension of the dynamics rather unreliable (e.g., \cite{principe1992prediction}). One reason for this fragility is that the embedded attractor in the reconstruction space not always an attractor of the map that is learnt in the reconstruction space, despite  unquestionably being an invariant set. When the embedded object is not explicitly known to be an attractor (as explained in Chapter \ref{ch3}) it can cause predictions to fail.

@@

This thesis deals with the implementation and analysis of non-interpretable models that can guarantee exact reconstruction. The project work concerns the study and implementation of a method (\cite{manjunath2021universal}) that incorporates learning a function by mapping the data on to a higher dimensional space using what is called a driven dynamical system (See Chapter 4). With a clear understanding of how the data is mapped onto the higher dimensional space, the method then permits the learning of a dynamical system topologically conjugate to that of the underlying system. Instead of linear regression as in the training of echo state networks, deep learning methods are employed to learn the correct function. With slight modifications to the implementation in the paper (with Adriaan), we show that one can construct accurate non-interpretable models with the ability to reconstruct attractors from more hard-to-forecast systems like the double pendulum. (The forecasting of the time-series from a double pendulum has not been reported before.) Moreover, we also demonstrate that long-term statistical consistency is preserved.

\emph{By solidifying the mathematical underpinnings of our theory, we hope to guarantee the ability to construct models with predictive power ranging from molecular biology to neuroscience. in the near future. (We can modify this sentence when the report is completed.)}

The more intricate details of proofs are referred to where relevant throughout. This report is organised as follows: 
\newline In Chapter~\ref{ch2}, we recall the definition of a discrete-time dynamical system, how a discrete-system arises from a flow of an ODE and then proceed to define the inverse limit space and topological conjugacy of autonomous systems. 
\newline In Chapter~\ref{ch3}, we introduce the problem of forecasting dynamical systems, state the Takens delay embedding theorem and discuss various issues faced while forecasting. 
\newline In Chapter~\ref{ch4}, a driven dynamical system is defined and discuss the properties of a specific class of driven dynamical systems that we make use of in this thesis/report/paper. 
\newline Finally in Chapter~\ref{ch5} we show the implementation of these forecasting methods, and conclusions are provided in Chapter~\ref{ch6}.

\chapter{Discrete-time Dynamical Systems}\label{ch2}

@@

Of course, discrete-time dynamical systems need not always arise through a differential equation, and once again we may consider the field of ecology where discrete dynamical systems are often directly derived or assumed. \textbf{Citation}
There is a school of thought that advocates discrete-time dynamical systems to be more natural for modelling real-world observations than differential equations and we refer the interested reader to \cite{saber2010introduction}.

@@

% \ednote{For let $B$ be an attractor of U. From above, we know that B is itself invariant, is asymptotically stable and has Lyapunov stability. Now suppose there is some attractor $C\subsetneq{U}=B$, so there is a $b\in{U}$ which is not in $C$. ........  M: Have changed the example}

@@

    \ednote{B: Is it desirable to have a more rigourous explanation here? }

@@

\ednote{M: Saying $\phi$ is then a semi-conjugacy between $T$ and $S$ could be ambiguous as it does not specify which system is on the top in the commutativity. So please edit accordingly. B: Does defining explicitly the domain and codomain of $\phi$ make it clear enough? }
When $S$ is conjugate to $T$, the dynamics of the two systems are in some way ''dynamically equivalent''. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$ \cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is also common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (e.g., \cite{de2013elements}) is a larger system capturing all of the important dynamics of its factor.

@@

First we establish the method of Takens delay embedding. 
In this method, one considers a discrete-time dynamical system defined on a 'nice' space (a smooth manifold)\ednote{Should this be further explained?} that can be obtained as time-$K$ map of a continuous time-dynamical system, concepts which will be formalised below.
We make observations from such a system, i.e., we observe the evolution of $\theta(w_0)$ (where $w_0$ represents the initial state) i.e., we observe  a finite set of values of $u_n :=  \theta(w_n) = \theta(Tw_{n-1})$. The sequence $\{u_n\}$ represents a scalar time-series and intuitively $\theta$ may be thought to represent a probe inserted into a larger system which is itself only measuring/extracting a small part of the greater system state at time $t=n$. 
Consider for example a thermometer erected to measure the ambient temperature in a local village. This measurement function, the thermometer, is capturing only a single aspect, the temperature, of a much grander dynamical system entailing the present weather of the surrounding area. Even more than that though- it is measuring a miniscule part of the global weather system.

@@

%Some explanation is beneficial. Takens establishes a delay-coordinate map $\Ftheta$ defined in \ref{eqn_takens} for $T$ the flow (\ref{defn_flow}) defined by the update equation  $Tu_n=u_{n+1}$ and $theta$ our measurement function. When confining a dynamical system to the manifold U (in which the underlying system is contained), Takens showed that if certain smoothness conditions are satisfied on $T$ and $\theta$, then the delay-coordinate map $\Ftheta$ embeds $U$ in the reconstruction space $R^{2d+1}$ \textbf{we haven't used the term reconstruction space yet, so this is falling out of the blue}, for almost every choice of measurement function $\theta$, the observable. By almost every choice here, we mean that we exclude some observations which do not give any information. To see a more precise discussion, see \textbf{xxx} for we do not concern ourselves with a further elaboration on this report.

@@

\begin{figure}[ht]
  \includegraphics[scale=0.25]{_takensmap.png}
  \centering
  \captionof{figure}{Schematic of Takens Embedding Theorem's usage and consequence of how iterates \ldots }
  \label{fig:takensmap}
\end{figure}
\ednote{M: Here, and throughout if you reproduce a figure, it must be stated where it was reproduced from. B: Is this satisfactory? Also, should $\tilde{F}$ be left out for now?}
\ednote{B: This is only a screenshot used as a temporary placeholder. If sufficient, I shall convert it to .eps. If not, may I perhaps have the original picture?}

@@

%Thus information about U can be retained in the time series' (or observation's) output. By preserving the topology on the manifold $U$ in the reconstruction space $X$ (\textbf{not discussed before}), we are also guaranteed the preservation of topological invariants of the manifold, of which dimensionality is one such invariant. (We note here that dimensionality will again be considered later in this project).

@@

It is a fact that Takens' Embedding Theorem is both a powerful result and that it provides compelling reason to believe that one could conceivably reconstruct accurately a system conjugate to the underlying system. Nonetheless, it does present some serious practical limitations:
\vspace{-5mm}
\begin{enumerate}
\item Even supposing that we could indeed find $F_\theta$ , our \emph{approximation} of  $F_\theta$ is a map from a larger set $\mathbb{R}^{2d+1}$ containing the embedded attractor. There are, however, no theoretical guarantees that $F_\theta$ will retain  $V=\Phi_{2d,\theta}(W)$  as an attractor although $W$ could itself be an attractor.
\item Takens theorem is stated only for noiseless observations. Due to noise $\epsilon_n$ the delay vector $\Phi_{2d,\theta}(w_n) + \epsilon_n$  may lie outside $V$.Furthermore, due to the chaotic nature of the underlying system (i.e. the fact that it has SIDC), the evolution of $\Phi_{2d,\theta}(w_n) + \epsilon_n$ under the map $F_\theta$ could move out of $V$ completely. This problem can be overcome  by using a driven dynamical system with some properties, and we discuss this in Chapter\ref{ch3}. 
\end{enumerate}

@@

The Takens embedding theorem does not guarantee global dissipativity.
\ednote{B:Links with~\ref{subs_LearnGamma} in discussion of dissipativity of $\Gamma$.  Sir, I'm not sure how to write this here - I think it a beneficial point to make so as to set the stage for the later discussion in ~\ref{subs_LearnGamma}, but I am unsure there as well.}

% \begin{figure}[ht]
%   \includegraphics[scale=0.3]{_learningfailure.png}
%   \centering
%   \caption{Illustration of a system having failed to learn the map}
%   \label{fig:supp_learning_failure}
% \end{figure}

@@

\section{Next section?}
\ednote{I feel a break is necessary here, but would hear from you what your thoughts are.}

@@

 \begin{Theorem}
  For a compactly driven system, a causal mapping $H$ exists if and only if $g$ has the USP. 
\end{Theorem}
\begin{proof}
  See~\cite[Th.3]{manjunath2013echo}.
\end{proof}

We direct our attention to the function \[H_2{\overline{}} := (h(r\overline{u}, h(\overline{u}))) \]

@@

The driven system $g$ can induce an embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ as follows: 
If $H:\overleftarrow{U}{\to}{\overleftarrow{X}_U}$ is also injective (in addition to being surjective, and hence bijective), 
it becomes the embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ induced by the driven system $g$ and we refer to $H$ as a \emph{causal embedding} 
\ednote{M : Causal embedding a dynamical system is different from the definition of causal embedding, so please edit/delete this paragraph and the next theorem and also in the remaining part till Section 1.2}.

@@

When $g$ has the USP, the diagram below illustrates the operation of the mappings $h$ and $H$. The mapping $h:\overleftarrow{U}\to{X_U}$ is an observable as discussed in an above section (\ref{TakensEmbedding}).

@@

\section{Choosing the driven system $g$}

@@

When a causal embedding $H$ exists for the driven system $g$, one can map an arbitrary input ${u}$ onto the solution space $X$ without additional distortion or information-loss \textbf{(Cite.)}.
When an embedding is established, the question of possible additional complexity in the solution is removed by guaranteeing that, since the systems are conjugate (semi-conjugate, \textbf{(refer)}), $g$ does not add any (some) complexity to the system.

@@

\begin{Definition}
  [\bf Uniform Attraction Property]\label{Dfn_UAP}\rm
  A driven system $g$ has the Uniform Attraction Property (UAP) if, regardless of starting position, all trajectories converge to a single trajectory as time flows forward. 
  This trajectory is also the unique solution sequence $x$ to the input sequence $u$ as mentioned above.
\end{Definition}

@@

\begin{figure}[ht]
  \includegraphics[scale=0.4]{_memloss_conttime.png}
  \centering
\captionof{figure}{Reproduced from \cite{manunath2021universal}. Illustration of all trajectories approaching the true solution if $g$ has the UAP}
\label{fig:memloss_conttime.png} 
\end{figure}

@@

This also solves the problem of perturbation/noise introduced by the observable or measurement function. Recall from before that an observable is inherently a map that discretises the underlying continuous-time system. Moreover, measurement and device error introduce mistakes. Since we are working with a chaotic system displaying sensitive dependence on initial conditions, these small errors could potentially send the trajectory into a completely different attractor. 
In the presence of the UAP, however, small measurement errors introduced into the system do not pose the same danger as before since the trajectories will eventually tend towards the actual solution.
\ednote{(B: Is this Complete enough?)}
% \textbf{From the perspective of perturbation of an autonomous dynamical system, if the resulting nonautonomous system has the USP, then the temporal structure of noise relates to the dynamics on a nonautonomous uniform attractor through a topological semi-conjugacy or a conjugacy.}

@@

% \ednote{Add Causal Embedding Theorem}
% \textbf{NB NB NB - Causal Embedding Theorem! - Th.4}

@@

% \begin{figure}[ht]
%   \includegraphics[scale=0.3]{_gammatemp.png}
%   \centering
%   \label{fig:temp_Gamma}
% \end{figure}

\begin{equation}  \label{Scomm_imp}
  %    \[ 
      \psset{arrows=->, arrowinset=0.25, linewidth=0.6pt, nodesep=3pt, labelsep=2pt, rowsep=1.3cm, colsep = 1.1cm, shortput =tablr}
   \everypsbox{\scriptstyle}
   \begin{psmatrix}
   \widehat{U}_T & \widehat{U}_T\\%
   Y_T & Y_T\\
   & \textcolor{red}{U \times X}
   %%%
  %  \ncline{1,1}{1,2}^{\widehat{T}} \ncline{1,1}{2,1} <{H_2}
  %  \ncline{1,2}{2,2} > {H_2}
  %  \ncline{2,1}{2,2}^{G_T}
  %  \psset{linecolor=red}
  %  \textcolor{red}{\ncline{2,1}{3,2}<{(\Gamma,\pi_2)}}
  %  \ncline{3,2}{2,2}>{\textcolor{red}{(\pi_2,g)}}.
   \end{psmatrix}
  % \]
  \end{equation}

@@

Informally, we may consider input-related stability as concerned with the question of whether or not small variations in input would result in small responses.   
If two inputs $u$ and $u_{noisy}$ are close in the product topology, then their tails could differ greatly due to the metric's insensitivity to the differences in sequence's tails.
 However if the system has the USP then the function $h$ is continuous and so $h(u), h(u_{noisy})$ remain near – maintaining a measure of robustness to input and measurement noise.\cite{manjunath2021universal}.
In a similar fashion, parameter-related stability is obtained thanks to a result \cite[Lemma 3.2]{manjunath2020stability} relating this form of stability with the ESP (which is here equivalent to the USP).

@@

\begin{figure}[ht]
  \includegraphics[scale=0.5]{_AutovsDriven.eps}
  \centering
  \label{fig:learningFailure}
  \captionof{figure}{Illustrating empirically the importance of global dissipativity. Plotted below the three principal components of randomly initiated trajectories of a learnt version of $G_T$ versus the trajectory $(x_{n-1}, x_n)$(grey) of driven stated obtained from the Thomas Attractor's data (see below). }
\end{figure}

@@

We make use of a Recurrent Neural Network (RNN) in our implementation. RNN's are widely used in the literature for discrete-time processing and are considered to be discrete-time version of continuous dynamical systems modelling a network of neurons. (\textbf{Citation.})
A complete discussion of RNN's is outside the scope of this thesis, so we present only the details necessary to the current discussion. For information, (\textbf{Cite, Cite, Cite}) may be consulted. We implement our RNN in the \emph{Python} programming language and making use of the \emph{Scipy} and \emph{Tensorflow} libraries.

If the RNN we will be using in the next step has $N$ neurons, then $A$ and $B$ are each $N\times{N}$ matrices. Inputs $u_n$ of dimension $K < N$ are padded by zeroes to obtain an input of dimension $N$. As we do not need the entire left-infinite history of input, an arbitrary initialisation of the system with initial values followed by the driving of the system with the function $g$ means that the generated values will approach the actual solution’s elements in a uniform manner thanks to the UAP. \textbf{(Citation)}

@@

\section{Changes in this project to previous work of~\cite{manjunath2021universal}}

We made several changes which gave improved results:
\vspace{-8mm}
\begin{enumerate}
\item Using Delay Coordinates (DC) when driving the system vs. using Single/Multi Delay Dynamics (SDD/MDD) after driving (and before training). Explain/show why these two methods are equivalent. \ednote{Would MDD  be not computationally intensive ? So may be you just explain if you can’t show it.}
\item Sparse matrix implementation. Slight improvement in some scenarios, but not fantastic. \ednote{M: Did it not help you a great deal with the DP forecasting? B: Sparse matrices tended to be finicky. In some cases they helped whereas in others they actually worsened the solution somewhat. There's some tuning to be done in terms of condition number, scaling and density}

(Possible paragraph here/later on implementing sparsity of A/B: Generate sparse matrix that is also well-conditioned and correctly scaled matrix.)
\item Density of matrices A/B(i.e. number of non-zero entries in each matrix) (separate from sparse matrix implementation)
\end{enumerate}

@@

The Double Pendulum consists of two pendulums fastened to one another such that the system moves as a whole. The first (or higher pendulum) is attached to a fixed point while the second is attached to the endpoint of the first. (See below)
This system is a classical example of a system exhibiting chaos. \textbf{(Citation.)} Consider again~\ref{fig:dp_sdic}.
The system has four variables: angle and angular velocity of each pendulum rod. An ideal system is assumed where the rods are massless and no friction in the pivots are considered.

\begin{figure}[ht]
  \includegraphics[scale=0.35]{_dp_setup.png}
  \centering
  \captionof{figure}{Pictorial representation of the setup of a DP. Image taken from~\cite{DPSetup}}
  \label{fig:dp_setup}
\end{figure}
